# Interactive Network Visualization

```{r, warning = FALSE, message = FALSE}
library(igraph)
library(threejs) # 3D Networks
library(visNetwork) # Interactive
library(htmlwidgets) 
library(htmltools)
library(dplyr)
```

In this chapter we will learn about how to create engaging network visualisations. By engaging, we mean so literally. One way to help your audience understand a network is to have them interact with it. R offers multiple different packages all filled with fun ways to accomplish this. In this chapter, we cover 2 methods one that creates 3D networks and a second that creates interactive networks.

## The Data

For this chapter, we are going to work with a Hollywood movie network. This is an affiliation network that contains information on the ten most popular movies (based on IMDB activity) for each year from 1999 to 2014. For each movie, the first ten actors listed under the IMDB field "Full Cast" are included. Thus, the underlying network data form a bipartite graph, where ties exist between a movie and the top ten actors who starred in that movie.

Note that it is not necessary to use two mode network data for either of the types of visualisations that we cover here. We simply use it to demonstrate that these methods can be used

First, we are going to read in out data and convert it into a network two mode network object. We will follow the basic procedure for mapping the two mode network and creating the vertex attribute denoting which type of node it is, an individual/actor (type = FALSE) or the group/movie (type = TRUE).

```{r}
load("Q:/Shared drives/SAT - The Three Amigos/SNA_BOOK_PACKAGES/intronets/inst/extdata/hwd_movies.RData")

hwd_g <- graph_from_data_frame(movies, directed = F)

V(hwd_g)$type <- bipartite_mapping(hwd_g)$type
```

## 3D Networks with threejs

To create the 3D Network we use the grpahjs() function which simply converts your static igraph object into an engaging model. This allows you to scroll in and out to zoom and click and drag to move around the network. It places the network in a spherical shape in the plot window and allows you to move in, out and around it.

```{r}
graphjs(hwd_g)
```

Now, we can do a little more ere to make the visualisation cleaner and more useful. Like working with igraph, either we can set some attributes to the graph directly (like some color or shape etc.) or we can state the options for the alterations when using the graphjs function. To demonstrate this, we will do a little of both.

First, let's leave the network visualisation pretty basic and just play with some of the parameters of graphjs. In this, we are going to change the sizes of the nodes, add a title, and set the labels to show the names of the nodes. In this, as you hover over each node the name of the node will populate.

```{r}
graphjs(hwd_g,   
  vertex.size = 0.25,
  vertex.label = V(hwd_g)$name,
  bg = "black",
  main = "Hollywod Movie Network"
)
```

This is much cleaner than the original visualisation. The contrast between the black backgorund and the orange nodes really makes the network pop and reducing the size of the nodes allows you to see the structure much clearer. There is a bit more we can do, however. Since this is a two mode network, let's add some color and shape vertex attributes so our viewers can differentiate between the actors and the movies.

```{r}

V(hwd_g)$shape <- ifelse(V(hwd_g)$type == TRUE, "square", "circle")
V(hwd_g)$color <- ifelse(V(hwd_g)$type == TRUE, "maroon", "ivory")

hwd_g_3d <- graphjs(hwd_g,   
  vertex.size = 0.25,
  vertex.label = V(hwd_g)$name,
  bg = "black",
  main = "Hollywod Movie Network"
)

hwd_g_3d
```

These 3D networks are really fun and a great way to engage viewers with the network. Next, we cover interactive visualisations that are slightly different from the above 3D networks. There are perks to both methods that become apparent as you use them.

## Interactive Visualisation with VisNetwork

VisNetwork does not just convert your network object into an interactive visualisation. Rather, we need to build a specific dataset that VisNetwork can pull the edge and node attributes from. To do this, we use out edgelist and construct a node characteristics data frame from our edgelist. Since our network is a two mode network we build one for movies and one for actors then merge them into one node-level dataset.

```{r}
movie <- unique(movies$movie)
actor <- unique(movies$actor)
```

```{r}
nodes <- data.frame(
  id = movie,
  label = movie,
  group = "movie",
  color = "maroon"
)

nodes2 <- data.frame(
  id = actor,
  label = "",
  group = "actor",
  color = "green"
)

nodes_done <- rbind(nodes, nodes2)
```

Now this is done, we need to swap the names of the columns in our edgelist to from and to. VisNetwork requires that the edgelist clearly delineates direction even if the graph (like ours) is undirected.

```{r}
d_done <- movies %>%
  rename(from = actor,to = movie)
```

With this, we can construct our basic interactive network. The basic recipe for this is to use the visNetwork() function and call on the edgelist and node attribute dataset. With this visualisaiton, you can click on and drag nodes selecting them one-by-one.

```{r}
visNetwork(
    nodes_done,
    d_done)
```

There is a lot more that we can do with this visualisation, however. The chunk below creates a much neater visualisation that we can use. It looks like a lot, but this is really just a sequence of small tweaks layered onto a base plot. In short, this process goes from initialising the network, defining the layout, defining groups, improve readability, adding interactivity, to adding legend. Let's break it down by each option we add so you can get a picture of what it all does.

We start with visNetwork, where visNetwork(nodes_done, d_done, width = "100%", height = "800px", background = "white") initializes the network and sets the canvas size and background. Next, visIgraphLayout(layout = "layout_nicely") uses igraph to space nodes in a clean, readable way.

The two visGroups() calls define the two-mode structure—actors as dots and movies as squares—so the node types are visually distinct. Next, visNodes(font = list(size = 30)) simply increases label size to improve readability. visOptions(highlightNearest = TRUE, nodesIdSelection = FALSE) adds light interactivity by highlighting connections on click while removing the dropdown selector. visPhysics(enabled = FALSE) freezes the layout so nodes don’t keep moving. You can drag and move the nodes around to place them where you want. Finally, visLegend(...) builds a custom legend, letting you control exactly how each node type is represented and where the legend appears.

```{r, eval=knitr::is_html_output()}
visNetwork(
    nodes_done,
    d_done,
    width = "100%",
    height = "800px",
    background = "white") %>%
  visIgraphLayout(layout = "layout_nicely") %>%
  visGroups(
    groupname = "actor",
    shape = "dot") %>%
  visGroups(
    groupname = "movie",
    shape = "square") %>%
  visNodes(
    font = list(size = 30)) %>%
  visOptions(
    highlightNearest = TRUE,
    nodesIdSelection = FALSE) %>%
  visPhysics(
    enabled = FALSE) %>%
  visLegend(
    useGroups = FALSE,
    addNodes = list(
      list(
        label = "Movie",
        shape = "square",
        color = "maroon",
        font = list(size = 10)),
      list(
        label = "actor",
        shape = "dot",
        color = "green",
        font = list(size = 10))),
    width = 0.15,
    position = "right")
```

Now we have that base, we are going to add a few extra cool things to it! We can add some physics to the visualisation, This introduces some repulsion physics that generate a 'springy' effect to the graph as you move the nodes. Rather than being able to place nodes where you want, this new network causes nodes to bound back into place after you drag it.

```{r}
visNetwork(
    nodes_done,
    d_done,
    width = "100%",
    height = "800px",
    background = "white") %>%
  visGroups(
    groupname = "actor",
    shape = "dot") %>%
  visGroups(
    groupname = "movie",
    shape = "square") %>%
  visNodes(
    font = list(size = 30)) %>%
  visOptions(
    highlightNearest = TRUE,
    nodesIdSelection = TRUE) %>%
  visLegend(
    useGroups = FALSE,
    addNodes = list(
      list(
        label = "Movie",
        shape = "square",
        color = "maroon",
        font = list(size = 10)),
      list(
        label = "Actor",
        shape = "dot",
        color = "green",
        font = list(size = 10))),
    width = 0.15,
    position = "right") %>%
visPhysics(
    solver = "repulsion",
    repulsion = list(
      nodeDistance = 100,  
      centralGravity = 0.1,
      springLength = 200,
      springConstant = 0.02,
      damping = 0.09
    ),
    stabilization = list(enabled = TRUE, iterations = 100)
  )
```

Here is a slightly different alteration of the physics with a slightly different layout, and we can add some control buttons directly to the visualisation so that users can navigate the visualisation by clicking not dragging their mouse.

```{r}
visNetwork(
    nodes_done,
    d_done,
    width = "100%",
    height = "800px",
    background = "white") %>%
  visGroups(
    groupname = "actor",
    shape = "dot") %>%
  visGroups(
    groupname = "movie",
    shape = "square") %>%
  visNodes(
    font = list(size = 30)) %>%
  visOptions(
    highlightNearest = TRUE,
    nodesIdSelection = TRUE) %>%
  visLegend(
    useGroups = FALSE,
    addNodes = list(
      list(
        label = "Movie",
        shape = "square",
        color = "maroon",
        font = list(size = 10)),
      list(
        label = "Actor",
        shape = "dot",
        color = "green",
        font = list(size = 10))),
    width = 0.15,
    position = "right") %>%
visPhysics(
    solver = "forceAtlas2Based",
    forceAtlas2Based = list(
      gravitationalConstant = -50,
      centralGravity = 0.01,
      springLength = 200,
      springConstant = 0.05,
      damping = 0.4
    ),
    stabilization = list(enabled = TRUE)
  ) %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = FALSE)%>%
  visInteraction(navigationButtons = TRUE)
```

## Saving Advanced Network Visualisations

Since these types of visualisations are not static, we can't simply save them to a jpeg or png file. Instead, they need to be stored in a file that will open and permit the interactivity. As such, we save them to an .html file that you can open in your broswer.

Here we use a combination of functions from a couple of html-friendly packages. Basically, what this does is ensure that the visualisation can be opened in a browser, then saves it to your working directory as a html file and opens it in your default browser.

```{r,eval= FALSE}
hwd_g_3d <- browsable(hwd_g_3d)

saveWidget(hwd_g_3d, file="hwd_movies.html")
browseURL("hwd_movies.html")
```

## References

For more on the Hollywood data...

- 

For more on Interactive and 3D Networks...

- <https://bwlewis.github.io/rthreejs/>

- <https://cran.r-project.org/web/packages/visNetwork/vignettes/Introduction-to-visNetwork.html>
